Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-26 paper/case-study-mean-std.tex
squeezed the paper into the new format

\usepackage{siunitx}
\usepackage[color]{changebar}
\usepackage{enumitem}
\usepackage[lighttt]{lmodern}
\usepackage{graphicx}

\usepackage{listings}

\ttfamily
\DeclareFontShape{OT1}{lmtt}{m}{it}
     {<->sub*lmtt/m/sl}{}
\lstdefinestyle{default}{language=Anglican,basicstyle=\ttfamily, columns=flexible, showstringspaces=false}

\usepackage{url}
\usepackage{etoolbox}

@@

@@

Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
represent generative models as programs written in a specialized
language that provides syntax for the definition and conditioning of
random variables. The code for such models is generally concise,
modular, and easy to modify or extend. Typically inference can be
performed for any probabilistic program using one or more generic
inference techniques provided by the system back end, such as
Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
\cite{SDT14}, expectation propagation \cite{MWG+10}, and extensions
of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
These generic techniques may be less statistically efficient
than techniques tailored to a specific model. However,
probabilistic programming facilitates simpler implementations of
models making the inference inherently faster.
\jw{I'm not sure I understand how PP facilitates simpler implementations that make inference inherently faster.}

While probabilistic programming systems shorten the iteration cycle in
exploratory model design, they typically lack basic functionality needed for
data I/O, pre-processing, and analysis and visualization of inference results.
In this paper, we describe the implementation of
Anglican~\cite{TMW15, Anglican}, a probabilistic
programming language that tightly integrates with
Clojure~\cite{H08,Clojure}, a general-purpose programming language
that runs on the Java Virtual Machine (JVM). Both languages share a common
syntax, and can be invoked from each other. This allows Anglican programs to
make use of a rich set of libraries written in both Clojure and Java.
Conversely, Anglican allows intuitive and compact specification of models for
which inference may be performed as part of a larger Clojure project.

@@

An Anglican program, or \textit{query}, is compiled into a
Clojure function. When inference is performed with a provided
algorithm, this produces a sequence of samples. Anglican shares
a common syntax with Clojure; Clojure functions can be called
from Anglican code and vice versa. A simple program in Anglican
can look like the following code:
%\begin{minipage}{\linewidth}
\begin{lstlisting}[style=default]
(defquery model "model selection" data
  (let [;;; Guess a distribution.
        dist (sample (categorical
                       [[normal 0.5]
                        [gamma 0.5]]))
        a (sample (gamma 1 1))
        b (sample (gamma 1 1))
        d (dist a b)]

    ;;; Observe samples from the distribution.
    (loop [observations data]
      (when (not-empty observations)
        ;; Retrieve the first observation as `o'
        (let [o (first observations)]
          ;; Observe 'o' from the guessed 
          ;; distribution 'd'.
          (observe d o))
        ;; Proceed to the next iteration with
        ;; the rest of observations.
        (recur (rest observations))))

    ;;; Return the distribution and parameters.
    [d a b]))
\end{lstlisting}

@@

Imagine that we are facing the following dilemma:
\begin{quote}
A customer wearing round sunglasses came to a deli at 1:13pm,
and grabbed a sandwich and a coffee. Later on the same day, a
customer wearing round sunglasses came at 6:09pm and ordered a
dinner.  Was it the same customer?
\end{quote}
Additionally, we know that:
\begin{itemize}
    \item There is an adjacent office quarter, and it takes
        between 5 and 15 minutes on average to walk from an
        office to the deli, where different average times are  for
        different buildings in the quarter.
    \item Depending on traffic lights, the walking time varies by about 2
        minutes.
    \item The lunch break starts at 1:00pm, and the workday ends at 6:00pm.
    \item Based on the similarity of appearance, the waiter assesses the odds that this is the same customer as 2 to 1.
\end{itemize}

@@

Besides computing the posterior probability that the same
customer visited the deli twice, we may want to know the average
time (or times) to arrive. In Bayesian inference, it is common
to report distributions instead of `most likely' values. We use
query results in retrieved samples to approximate the
distributions, and plot distribution histograms for the same
customer visiting twice (Figure~\ref{fig:dist-same}) and
two different customers (Figure~\ref{fig:dist-diff})

@@

;; invoked when algo parameter is :importance
(defmethod infer :importance 
  [algo prog value & {}]
  (letfn
    ;; recursive function without any parameter
    [sample-seq 
      []
      ;; lazy infinite sequence 
      (lazy-seq
        (cons
          (:state (exec ::algorithm prog value
                        initial-state))
          (sample-seq))))]
    (sample-seq)))
\end{lstlisting}
This implementation is dispatched 
when \texttt{infer} receives {\tt :im\-por\-tan\-ce} as its \texttt{algo} 
parameter. (Anglican includes multiple implementations of \texttt{infer}
for different inference algorithms.) Once dispatched, it lazily 
constructs an infinite sequence of inference states by repeatedly 
executing the program \texttt{prog} using \texttt{exec}
and retrieving the final inference state of the execution 
using \texttt{:state}. For checkpoint handlers, importance sampling simply
relies on their default implementations.

>>>>

\usepackage{siunitx}
\usepackage[color]{changebar}
\usepackage{enumitem}
\usepackage[lighttt]{lmodern}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{url}
\usepackage{etoolbox}

\ttfamily
\DeclareFontShape{OT1}{lmtt}{m}{it}
     {<->sub*lmtt/m/sl}{}
\iftoggle{full}{
\lstdefinestyle{default}{language=Anglican,basicstyle=\ttfamily\small, columns=flexible, showstringspaces=false}
}{
\lstdefinestyle{default}{language=Anglican,basicstyle=\linespread{0.8}\ttfamily\small, columns=flexible, showstringspaces=false}
}

@@

\begin{CCSXML}
<ccs2012>
    <concept>
        <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
        <concept_desc>Software and its engineering~Functional languages</concept_desc>
        <concept_significance>300</concept_significance>
    </concept>
    <concept>
        <concept_id>10011007.10011006.10011050.10011023</concept_id>
        <concept_desc>Software and its engineering~Specialized application languages</concept_desc>
        <concept_significance>300</concept_significance>
    </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[300]{Software and its engineering~Functional languages}
\ccsdesc[300]{Software and its engineering~Specialized application languages}

@@

Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
represent generative models as programs written in a specialized
language that provides syntax for the definition and conditioning of
random variables. The code for such models is generally concise,
modular, and easy to modify or extend. Typically inference can be
performed for any probabilistic program using one or more generic
inference techniques provided by the system back end, such as
Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
\cite{SDT14}, expectation propagation \cite{MWG+10}, and extensions
of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.

While probabilistic programming systems shorten the iteration cycle in
exploratory model design, they typically lack basic functionality needed for
data I/O, pre-processing, and analysis and visualization of inference results.
In this paper, we describe the implementation of
Anglican~\cite{TMW15, Anglican}, a probabilistic
programming language that tightly integrates with
Clojure~\cite{H08}, a general-purpose programming language
that runs on the Java Virtual Machine (JVM). Both languages share a common
syntax, and can be invoked from each other. This allows Anglican programs to
make use of a rich set of libraries written in both Clojure and Java.
Conversely, Anglican allows intuitive and compact specification of models for
which inference may be performed as part of a larger Clojure project.

@@

An Anglican program, or \textit{query}, is compiled into a
Clojure function. When inference is performed with a provided
algorithm, this produces a sequence of samples. Anglican shares
a common syntax with Clojure; Clojure functions can be called
from Anglican code and vice versa. A simple program in Anglican
can look like the following code:
%\begin{minipage}{\linewidth}
\begin{lstlisting}[style=default]
(defquery model "model selection" data
  (let [;;; Guess a distribution.
        dist (sample (categorical
                       [[normal 0.5]
                        [gamma 0.5]]))
        a (sample (gamma 1 1))
        b (sample (gamma 1 1))
        d (dist a b)]
    ;;; Observe samples from the distribution.
    (loop [observations data]
      (when (not-empty observations)
        ;; Retrieve the first observation as `o'
        (let [o (first observations)]
          ;; Observe 'o' from the guessed 
          ;; distribution 'd'.
          (observe d o))
        ;; Proceed to the next iteration with
        ;; the rest of observations.
        (recur (rest observations))))
    ;;; Return the distribution and parameters.
    [d a b]))
\end{lstlisting}

@@

Imagine that we are facing the following dilemma:

A customer wearing round sunglasses came to a deli at 1:13pm,
and grabbed a sandwich and a coffee. Later on the same day, a
customer wearing round sunglasses came at 6:09pm and ordered a
dinner.  Was it the same customer?

Additionally, we know that:
\begin{itemize}
    \item There is an adjacent office quarter, and it takes
        between 5 and 15 minutes on average to walk from an
        office to the deli, where different average times are  for
        different buildings in the quarter.
    \item Depending on traffic lights, the walking time varies by about 2
        minutes.
    \item The lunch break starts at 1:00pm, and the workday ends at 6:00pm.
    \item Based on the similarity of appearance, the waiter assesses the odds that this is the same customer as 2 to 1.
\end{itemize}

@@

Besides computing the posterior probability that the same
customer visited the deli twice, we may want to know the average
time (or times) to arrive. In Bayesian inference, it is common
to report distributions instead of `most likely' values. We use
query results in retrieved samples to approximate the
distributions, and plot distribution histograms for the same
customer visiting twice (Figure~\ref{fig:dist-same}) and
two different customers (Figure~\ref{fig:dist-diff}).

@@

;; invoked when algo parameter is :importance
(defmethod infer :importance 
  [algo prog value & {}]
  (letfn
    ;; recursive function without any parameter
    [sample-seq []
      ;; lazy infinite sequence 
      (lazy-seq (cons
                   (:state (exec ::algorithm prog value
                                   initial-state))
                   (sample-seq)))]
    (sample-seq)))
\end{lstlisting}
This implementation is dispatched 
when \texttt{infer} receives {\tt :im\-por\-tan\-ce} as its \texttt{algo} 
parameter. (Anglican includes multiple implementations of \texttt{infer}
for different inference algorithms.) Once dispatched, it lazily 
constructs an infinite sequence of inference states by repeatedly 
executing the program \texttt{prog} using \texttt{exec}
and retrieving the final inference state of the execution 
using \texttt{:state}. For checkpoint handlers, importance sampling simply
relies on their default implementations.
